---
title: "index"
date: 02-09-2026
author: Loïc Thurre
format:
    html:
      toc: true
      toc-location: left

  
---

# 16S Root bacterial profiling with PacBio + Kinnex
This is the analysis that orginates from the following [lab book](https://wiki.biozentrum.unibas.ch/spaces/DUWPMI/pages/574286083/20251110+-+LTh+FF+-+2nd+trial+-+Protocol+16S+profiling+of+plant+roots+with+Kinnex-PacBio).

**Overall goal**: develop a method that uses PacBio + Kinnex to profile the root bacterial communities. 

**Specific goal**: compare two primer pairs with Kinnex adapters:

1. **27F - 1492R** = 16S long-read. PacBio Kinnex documentation
2. **27F - 1185mR** = 16S fragment, excluding plastids (Lutzoni lab). Custom Kinnex adapters



I separate the analysis by SampleOwner and filtered/unfiltered data. 


```{r}
#| label: package import
#| echo: false
#| warning: false

library(dplyr)
library(ggplot2)
library(phyloseq)
library(tidyr)
library(tibble)
library(openxlsx)
```

```{r}
#| label: utils
#| warning: false

 quickRareCurve <- function (x, nsteps = 50, max.cores = T, nCores = 1, ...){
  require(parallel)
  require(vegan)
  x <- as.matrix(x)
  if (!identical(all.equal(x, round(x)), TRUE)){stop("function accepts only integers (counts)")}
  tot <- rowSums(x) # calculates library sizes
  S <- specnumber(x) # calculates n species for each sample
  if (any(S <= 0)) {
    message("empty rows removed")
    x <- x[S > 0, , drop = FALSE]
    tot <- tot[S > 0]
    S <- S[S > 0]
  } # removes any empty rows
  nr <- nrow(x) # number of samples
  # parallel mclapply
  # set number of cores
  mc <- getOption("mc.cores", ifelse(max.cores, detectCores(), nCores))
  message(paste("Using ", mc, " cores"))
  out <- mclapply(seq_len(nr), mc.cores = mc, function(i) {
    #different then in normal rarecurve function, here we calculate stepsize so that we have n steps per sample
    stepsize <- max(as.integer(tot[i]/(nsteps)), 1)
    n <- seq(1, tot[i], by = stepsize)
    if (n[length(n)] != tot[i]){n <- c(n, tot[i])}
    drop(rarefy(x[i, ], n))
  })
  #add names
  names(out) <- rownames(x)
  #long-transformation
  protox <- mapply(FUN = function(x, y) {
    mydf <- as.data.frame(x)
    colnames(mydf) <- "value"
    mydf$sample_name <- y
    mydf$subsample <- attr(x, "Subsample")
    mydf
  }, x = out, y = as.list(names(out)), SIMPLIFY = FALSE)
  out_long <- do.call(rbind, protox)
  rownames(out_long) <- NULL  # pretty
   
  return(out_long)
 }
 
 plot_theme = theme(
   axis.text = element_text(size = 12),
   axis.title = element_text(size = 12),
   legend.title = element_text(size = 14, face = "bold"),
   legend.text = element_text(size = 12),
   strip.text = element_text(size = 14)
 )
```

# 97% Clustering
## 0. Data import

```{r}
#| label: import data 1492R
#| echo: false
#| warning: false
  
  # DESIGN
  design.1492 <- read.xlsx("../00_DATA/primer_list_fiona_loic.xlsx", sheet = 1) %>% 
    filter(PrimerCombo == "27-1492") %>% 
    select(c(Sample, PrimerComb, SampleOwner, extraction_method, host)) %>% 
    tibble::column_to_rownames("PrimerComb")

  # OTU table
  otu.1492 <- read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_COUNT97.tab")
  rownames(otu.1492) = gsub("PB403rev1492.", "", rownames(otu.1492))
  rownames(otu.1492) = gsub("_cutted_filt", "", rownames(otu.1492))
  
  otu.sample.1492 <- merge(design.1492 %>% dplyr::select(Sample), otu.1492, by = 0) %>% 
    tibble::column_to_rownames("Sample") 
  
  master.1492 <- merge(design.1492, otu.1492, by = 0) 
  
  # tax_table
  tax.1492 <- data.frame(read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_TAXA97.tab"))
  
  tax.bac.1492 <- read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
    filter(Kingdom == "Bacteria") %>% 
    filter(Family != "Mitochondria") %>% 
    filter(Order != "Chloroplast")
  
  tax.plant.1492 <- read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
    filter(Kingdom != "Bacteria")
  

  
  physeq.1492 <- phyloseq(otu_table(otu.1492, taxa_are_rows = FALSE), sample_data(design.1492),   tax_table(as.matrix(tax.1492)))
```

```{r}
#| label: import data 1185
#| echo: false
#| warning: false
  
  # DESIGN
  design.1185 <- read.xlsx("../00_DATA/primer_list_fiona_loic.xlsx", sheet = 1) %>% 
    filter(PrimerCombo == "27-1492") %>% 
    select(c(Sample, PrimerComb, SampleOwner, extraction_method, host)) %>% 
    tibble::column_to_rownames("PrimerComb")

  # OTU table
  otu.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_COUNT97.tab")
  rownames(otu.1185) = gsub("PB403rev1185.", "", rownames(otu.1185))
  rownames(otu.1185) = gsub("_cutted_filt", "", rownames(otu.1185))
  
  otu.sample.1185 <- merge(design.1185 %>% dplyr::select(Sample), otu.1185, by = 0) %>% 
    tibble::column_to_rownames("Sample") 
  
  master.1185 <- merge(design.1185, otu.1185, by = 0) 
  
  # tax_table
  tax.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_TAXA97.tab")
    tax.bac.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
      filter(Kingdom == "Bacteria") %>% 
      filter(Family != "Mitochondria") %>% 
      filter(Order != "Chloroplast")
  
  tax.plant.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
    filter(Kingdom != "Bacteria")
  
  physeq.1185 <- phyloseq(otu_table(otu.1185, taxa_are_rows = FALSE), sample_data(design.1185), tax_table(as.matrix(tax.1185)))
```


## 1. Fiona
### 1.1 Unfiltered data 

Unfiltered data: all OTUs are kept. Including the plant, mitochondria and plastids.

#### 1.1.1 1492R

```{r}
#| echo: false
#| warning: false

  fiona.1492 <- master.1492 %>% 
    filter(SampleOwner == "Fiona")
  
  analyse.fiona.1492 <- fiona.1492 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1492), "BAC", "PLANT"))


  analyse.fiona.1492 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
  

  analyse.fiona.1492 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = extraction_method, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1492R") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
    
      

  analyse.fiona.1492 %>% 
    group_by(Sample, SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC + PLANT, fill = extraction_method)) +
    geom_col() + ylab("Number of reads per sample") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1492R") + theme(axis.text.x = element_text(angle = 90)) +
    plot_theme
  
  analyse.fiona.1492 %>% 
    group_by(Sample, extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1492R") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
```

#### 1.1.2 1185mR

```{r}
#| echo: false
#| warning: false

  fiona.1185 <- master.1185 %>% 
    filter(SampleOwner == "Fiona")

  analyse.fiona.1185 <- fiona.1185 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1185), "BAC", "PLANT"))
  
    analyse.fiona.1185 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
  

  analyse.fiona.1185 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = extraction_method, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1185mR") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
    
      

  analyse.fiona.1185 %>% 
    group_by(Sample, SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC + PLANT, fill = extraction_method)) +
    geom_col() + ylab("Number of reads per sample") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1185mR") + theme(axis.text.x = element_text(angle = 90)) +
    plot_theme
  
  
  
  analyse.fiona.1185 %>% 
    group_by(Sample, extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1185mR") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
```
#### 1.1.3 Compare 1185mR with 1492R


```{r}
#| echo: false
#| warning: false
  
  summary.fiona.1185 <- analyse.fiona.1185 %>% 
    mutate(primers = "27F_1185mR") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    )


   summary.fiona.1492 <- analyse.fiona.1492 %>% 
    mutate(primers = "27F_1492R") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    )
   
   
   summary.fiona <- rbind(summary.fiona.1185, summary.fiona.1492)
   
   
   
   ggplot(summary.fiona, aes(x = Sample, y = BAC /(BAC + PLANT), fill = primers)) +
    geom_col(position = "dodge") + ylab("Proportion of bacteria per read") +
    theme_bw() +
    facet_wrap(~host * extraction_method) +
    theme(axis.text.x = element_text(angle = 90)) + plot_theme
   
   ggplot(summary.fiona, aes(x = primers, y = BAC / (BAC + PLANT), label = Sample, fill = primers)) +
     geom_col() +
     facet_wrap(~ Sample ) + 
     theme_bw() + plot_theme + theme(axis.text.x = element_blank())
```


```{r}
#| echo: false
#| warning: false

  # rarefaction curves
  rar.fiona.1492 <- quickRareCurve(otu.sample.1492[11:nrow(otu.sample.1492),-1])
  
  ggplot(rar.fiona.1492, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1492R, Maize samples / Unfiltered") +
    plot_theme
  
  
  
  rar.fiona.1185 <- quickRareCurve(otu.sample.1185[11:nrow(otu.sample.1492),-1])
  
  ggplot(rar.fiona.1185, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1185mR, Maize samples") +
    plot_theme
```

The rarefaction curves show a higher diversity for the 1185mR primer pair. However, it seems fishy as it is in only one sample.

Let's have a look the primer pair is capturing one family or phylum that is then dominating.

```{r}
#| fig-height: 12
#| fig-width: 10
  physeq.1185.phylum <- tax_glom(physeq.1185, taxrank = "Phylum")

  physeq.1185.phylum.rel <- transform_sample_counts(
    physeq.1185.phylum,
    function(x) x / sum(x)
  )

  percentages_1185.rel.fiona <- psmelt(physeq.1185.phylum.rel) %>% filter(SampleOwner == "Fiona")
  
  
  relative_plot <- ggplot(data=percentages_1185.rel.fiona, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1185mR") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
  
  
  physeq.1492.phylum <- tax_glom(physeq.1492, taxrank = "Phylum")

  physeq.1492.phylum.rel <- transform_sample_counts(
    physeq.1492.phylum,
    function(x) x / sum(x)
  )

  percentages_1492.rel.fiona <- psmelt(physeq.1492.phylum.rel) %>% filter(SampleOwner == "Fiona")
  
  
  relative_plot <- ggplot(data=percentages_1492.rel.fiona, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1492R") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
```
Like this, it's hard to read and not very informative. But the 1185mR tax_table is dominated by one family (Aphagea_fa) which corresponds to a Eukaryote. Aside from this, the two tax tables are pretty similar. Although they are dominated by chloroplasts and mitochondria... 

If there is a difference, it will be much more clear in the filtered data part. 

### 1.2 Filtered data 

Filtered data, only strictly bacteria are kept. 

Rules based on taxonomy assignment:

    - filter(Kingdom == "Bacteria") 
    - filter(Family != "Mitochondria")
    - filter(Order != "Chloroplast")

#### 1.2.1 1492R

```{r}
#| echo: false
#| warning: false
  otu.1492.bac <- otu.1492 %>% dplyr::select(all_of(rownames(tax.bac.1492)))
  
  otu.sample.1492 <- merge(design.1492 %>% dplyr::select(Sample), otu.1492.bac, by = 0) %>% 
  tibble::column_to_rownames("Sample") %>% 
    filter()
  
  master.1492 <- merge(design.1492, otu.1492.bac, by = 0) 


  fiona.1492 <- master.1492 %>% 
    filter(SampleOwner == "Fiona")
  
  analyse.fiona.1492 <- fiona.1492 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1492), "BAC", "PLANT"))


  analyse.fiona.1492 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
```

#### 1.2.2 1185mR

```{r}
#| echo: false
#| warning: false

  otu.1185.bac <- otu.1185 %>% dplyr::select(all_of(rownames(tax.bac.1185)))
  
  otu.sample.1185 <- merge(design.1185 %>% dplyr::select(Sample), otu.1185.bac, by = 0) %>% 
  tibble::column_to_rownames("Sample") %>% 
    filter()
  
  master.1185 <- merge(design.1185, otu.1185.bac, by = 0) 

  fiona.1185 <- master.1185 %>% 
    filter(SampleOwner == "Fiona")

  analyse.fiona.1185 <- fiona.1185 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1185), "BAC", "PLANT"))
  
    analyse.fiona.1185 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
```
#### 1.2.3 Compare 1185mR with 1492R


```{r}
#| echo: false
#| warning: false
  
  summary.fiona.1185 <- analyse.fiona.1185 %>% 
    mutate(primers = "27F_1185mR") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value))


   summary.fiona.1492 <- analyse.fiona.1492 %>% 
    mutate(primers = "27F_1492R") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value))
   
   
   summary.fiona <- rbind(summary.fiona.1185, summary.fiona.1492)
   
   
   
   
   ggplot(summary.fiona, aes(x = primers, y = OTUsum, label = Sample, fill = primers)) +
     geom_col() +
     facet_wrap(~ Sample ) + 
     theme_bw() + plot_theme + theme(axis.text.x = element_blank()) + ylab("Number of bacterial reads")
```


```{r}
#| echo: false
#| warning: false

  # rarefaction curves
  rar.fiona.1492 <- quickRareCurve(otu.sample.1492[11:nrow(otu.sample.1492),-1])
  
  ggplot(rar.fiona.1492, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1492R, Maize samples / Unfiltered") +
    plot_theme
  
  
  
  rar.fiona.1185 <- quickRareCurve(otu.sample.1185[11:nrow(otu.sample.1492),-1])
  
  ggplot(rar.fiona.1185, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1185mR, Maize samples") +
    plot_theme
```

The rarefaction curves show a higher diversity for the 1185mR primer pair. 

```{r}
#| fig-height: 12
#| fig-width: 10

  physeq.1185.filtered <- prune_taxa(rownames(tax.bac.1185), physeq.1185)
  physeq.1185.phylum <- tax_glom(physeq.1185.filtered, taxrank = "Phylum")

  physeq.1185.phylum.rel <- transform_sample_counts(
    physeq.1185.phylum,
    function(x) x / sum(x)
  )

  percentages_1185.rel.fiona <- psmelt(physeq.1185.phylum.rel) %>% filter(SampleOwner == "Fiona")
  
  
  relative_plot <- ggplot(data=percentages_1185.rel.fiona, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1185mR") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
  
  physeq.1492.filtered <- prune_taxa(rownames(tax.bac.1492), physeq.1492)
  physeq.1492.phylum <- tax_glom(physeq.1492.filtered, taxrank = "Phylum")

  physeq.1492.phylum.rel <- transform_sample_counts(
    physeq.1492.phylum,
    function(x) x / sum(x)
  )

  percentages_1492.rel.fiona <- psmelt(physeq.1492.phylum.rel) %>% filter(SampleOwner == "Fiona")
  
  
  relative_plot <- ggplot(data=percentages_1492.rel.fiona, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1492R") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
```

## 2. Loïc 

### 2.1 Unfiltered data 

Unfiltered data: all OTUs are kept. Including the plant, mitochondria and plastids.

```{r}
#| label: reimport data 1492R
#| echo: false
#| warning: false
  
  # DESIGN
  design.1492 <- read.xlsx("../00_DATA/primer_list_fiona_loic.xlsx", sheet = 1) %>% 
    filter(PrimerCombo == "27-1492") %>% 
    select(c(Sample, PrimerComb, SampleOwner, extraction_method, host)) %>% 
    tibble::column_to_rownames("PrimerComb")

  # OTU table
  otu.1492 <- read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_COUNT97.tab")
  rownames(otu.1492) = gsub("PB403rev1492.", "", rownames(otu.1492))
  rownames(otu.1492) = gsub("_cutted_filt", "", rownames(otu.1492))
  
  otu.sample.1492 <- merge(design.1492 %>% dplyr::select(Sample), otu.1492, by = 0) %>% 
    tibble::column_to_rownames("Sample") 
  
  master.1492 <- merge(design.1492, otu.1492, by = 0) 
  
  # tax_table
  tax.1492 <- data.frame(read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_TAXA97.tab"))
  
  tax.bac.1492 <- read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
    filter(Kingdom == "Bacteria") %>% 
    filter(Family != "Mitochondria") %>% 
    filter(Order != "Chloroplast")
  
  tax.plant.1492 <- read.table("../00_DATA/R1492_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
    filter(Kingdom != "Bacteria")
  

  
  physeq.1492 <- phyloseq(otu_table(otu.1492, taxa_are_rows = FALSE), sample_data(design.1492),   tax_table(as.matrix(tax.1492)))
```

```{r}
#| label: reimport data 1185
#| echo: false
#| warning: false
  
  # DESIGN
  design.1185 <- read.xlsx("../00_DATA/primer_list_fiona_loic.xlsx", sheet = 1) %>% 
    filter(PrimerCombo == "27-1492") %>% 
    select(c(Sample, PrimerComb, SampleOwner, extraction_method, host)) %>% 
    tibble::column_to_rownames("PrimerComb")

  # OTU table
  otu.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_COUNT97.tab")
  rownames(otu.1185) = gsub("PB403rev1185.", "", rownames(otu.1185))
  rownames(otu.1185) = gsub("_cutted_filt", "", rownames(otu.1185))
  
  otu.sample.1185 <- merge(design.1185 %>% dplyr::select(Sample), otu.1185, by = 0) %>% 
    tibble::column_to_rownames("Sample") 
  
  master.1185 <- merge(design.1185, otu.1185, by = 0) 
  
  # tax_table
  tax.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_TAXA97.tab")
    tax.bac.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
      filter(Kingdom == "Bacteria") %>% 
      filter(Family != "Mitochondria") %>% 
      filter(Order != "Chloroplast")
  
  tax.plant.1185 <- read.table("../00_DATA/R1185_OUTPUT/bacteria/97/bacteria_TAXA97.tab") %>%
    filter(Kingdom != "Bacteria")
  
  physeq.1185 <- phyloseq(otu_table(otu.1185, taxa_are_rows = FALSE), sample_data(design.1185), tax_table(as.matrix(tax.1185)))
```

#### 2.1.1 1492R

```{r}
#| echo: false
#| warning: false

  loic.1492 <- master.1492 %>% 
    filter(SampleOwner == "Loïc")
  
  analyse.loic.1492 <- loic.1492 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1492), "BAC", "PLANT"))


  analyse.loic.1492 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
  

  analyse.loic.1492 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = extraction_method, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1492R") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
    
      

  analyse.loic.1492 %>% 
    group_by(Sample, SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC + PLANT, fill = extraction_method)) +
    geom_col() + ylab("Number of reads per sample") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1492R") + theme(axis.text.x = element_text(angle = 90)) +
    plot_theme
  
  analyse.loic.1492 %>% 
    group_by(Sample, extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1492R") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
```

#### 1.1.2 1185mR

```{r}
#| echo: false
#| warning: false

  loic.1185 <- master.1185 %>% 
    filter(SampleOwner == "Loïc")

  analyse.loic.1185 <- loic.1185 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1185), "BAC", "PLANT"))
  
    analyse.loic.1185 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
  

  analyse.loic.1185 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = extraction_method, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1185mR") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
    
      

  analyse.loic.1185 %>% 
    group_by(Sample, SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC + PLANT, fill = extraction_method)) +
    geom_col() + ylab("Number of reads per sample") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1185mR") + theme(axis.text.x = element_text(angle = 90)) +
    plot_theme
  
  
  
  analyse.loic.1185 %>% 
    group_by(Sample, extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    ) %>% 
  ggplot(., aes(x = Sample, y = BAC / (BAC + PLANT) * 100, fill = extraction_method)) +
    geom_col() + ylab("Proportion of bacterial reads (BAC / (BAC + PLANT) (%)") +
    theme_bw() +
    facet_wrap(~host) +
    ggtitle("27F 1185mR") +
    scale_y_continuous(breaks = c(0,25,50,75,100)) +
    plot_theme + theme(
      axis.text.x = element_text(angle = 90)
    )
```
#### 1.1.3 Compare 1185mR with 1492R


```{r}
#| echo: false
#| warning: false
  
  summary.loic.1185 <- analyse.loic.1185 %>% 
    mutate(primers = "27F_1185mR") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    )


   summary.loic.1492 <- analyse.loic.1492 %>% 
    mutate(primers = "27F_1492R") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value)) %>% 
      pivot_wider(
      names_from  = organism,
      values_from = OTUsum
    )
   
   
   summary.loic <- rbind(summary.loic.1185, summary.loic.1492)
   
   
   
   ggplot(summary.loic, aes(x = Sample, y = BAC /(BAC + PLANT), fill = primers)) +
    geom_col(position = "dodge") + ylab("Proportion of bacteria per read") +
    theme_bw() +
    facet_wrap(~host * extraction_method) +
    theme(axis.text.x = element_text(angle = 90)) + plot_theme
   
   ggplot(summary.loic, aes(x = primers, y = BAC / (BAC + PLANT), label = Sample, fill = primers)) +
     geom_col() +
     facet_wrap(~ Sample ) + 
     theme_bw() + plot_theme + theme(axis.text.x = element_blank())
```


```{r}
#| echo: false
#| warning: false

  # rarefaction curves
  rar.loic.1492 <- quickRareCurve(otu.sample.1492[1:10,-1])
  
  ggplot(rar.loic.1492, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1492R, Maize samples / Unfiltered") +
    plot_theme
  
  
  
  rar.loic.1185 <- quickRareCurve(otu.sample.1185[1:10,-1])
  
  ggplot(rar.loic.1185, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1185mR, Maize samples") +
    plot_theme
```

The rarefaction curves show a higher diversity for the 1185mR primer pair. However, it seems fishy as it is in only one sample.

Let's have a look the primer pair is capturing one family or phylum that is then dominating.

```{r}
#| fig-height: 12
#| fig-width: 10
  physeq.1185.phylum <- tax_glom(physeq.1185, taxrank = "Phylum")

  physeq.1185.phylum.rel <- transform_sample_counts(
    physeq.1185.phylum,
    function(x) x / sum(x)
  )

  percentages_1185.rel.loic <- psmelt(physeq.1185.phylum.rel) %>% filter(SampleOwner == "Loïc")
  
  
  relative_plot <- ggplot(data=percentages_1185.rel.loic, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1185mR") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
  
  
  physeq.1492.phylum <- tax_glom(physeq.1492, taxrank = "Phylum")

  physeq.1492.phylum.rel <- transform_sample_counts(
    physeq.1492.phylum,
    function(x) x / sum(x)
  )

  percentages_1492.rel.loic <- psmelt(physeq.1492.phylum.rel) %>% filter(SampleOwner == "Loïc")
  
  
  relative_plot <- ggplot(data=percentages_1492.rel.loic, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1492R") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
```


### 2.2 Filtered data 

Filtered data, only strictly bacteria are kept. 

Rules based on taxonomy assignment:

    - filter(Kingdom == "Bacteria") 
    - filter(Family != "Mitochondria")
    - filter(Order != "Chloroplast")

#### 2.2.1 1492R

```{r}
#| echo: false
#| warning: false
  otu.1492.bac <- otu.1492 %>% dplyr::select(all_of(rownames(tax.bac.1492)))
  
  otu.sample.1492 <- merge(design.1492 %>% dplyr::select(Sample), otu.1492.bac, by = 0) %>% 
  tibble::column_to_rownames("Sample") %>% 
    filter()
  
  master.1492 <- merge(design.1492, otu.1492.bac, by = 0) 


  loic.1492 <- master.1492 %>% 
    filter(SampleOwner == "Loïc")
  
  analyse.loic.1492 <- loic.1492 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1492), "BAC", "PLANT"))


  analyse.loic.1492 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
```

#### 2.2.2 1185mR

```{r}
#| echo: false
#| warning: false

  otu.1185.bac <- otu.1185 %>% dplyr::select(all_of(rownames(tax.bac.1185)))
  
  otu.sample.1185 <- merge(design.1185 %>% dplyr::select(Sample), otu.1185.bac, by = 0) %>% 
  tibble::column_to_rownames("Sample") %>% 
    filter()
  
  master.1185 <- merge(design.1185, otu.1185.bac, by = 0) 

  loic.1185 <- master.1185 %>% 
    filter(SampleOwner == "Loïc")

  analyse.loic.1185 <- loic.1185 %>% 
    pivot_longer(cols = starts_with("OTU")) %>% 
    mutate(organism = ifelse(name %in% rownames(tax.bac.1185), "BAC", "PLANT"))
  
    analyse.loic.1185 %>% 
    group_by(SampleOwner,extraction_method, host, organism) %>% 
    summarise(OTUsum = sum(value)) %>% 
    ggplot(., aes( x = organism, y = OTUsum, fill = organism)) +
      geom_col() +
      theme_bw() +
      facet_grid(extraction_method ~ host) +
      ylab("Summmed reads per organism") +
      plot_theme
```

#### 2.2.3 Compare 1185mR with 1492R

```{r}
#| echo: false
#| warning: false
  
  summary.loic.1185 <- analyse.loic.1185 %>% 
    mutate(primers = "27F_1185mR") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value))


   summary.loic.1492 <- analyse.loic.1492 %>% 
    mutate(primers = "27F_1492R") %>% 
    group_by(Sample,extraction_method, host, organism, primers) %>% 
    summarise(OTUsum = sum(value))
   
   
   summary.loic<- rbind(summary.loic.1185, summary.loic.1492)
   
   
   
   
   ggplot(summary.loic, aes(x = primers, y = OTUsum, label = Sample, fill = primers)) +
     geom_col() +
     facet_wrap(~ Sample ) + 
     theme_bw() + plot_theme + theme(axis.text.x = element_blank()) + ylab("Number of bacterial reads")
```


```{r}
#| echo: false
#| warning: false

  # rarefaction curves
  rar.loic.1492 <- quickRareCurve(otu.sample.1492[1:10,-1])
  
  ggplot(rar.loic.1492, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1492R, Maize samples / Unfiltered") +
    plot_theme
  
  
  
  rar.loic.1185 <- quickRareCurve(otu.sample.1185[1:10,-1])
  
  ggplot(rar.loic.1185, aes(x = subsample, y = value, group = sample_name, color = sample_name)) +
    theme_bw() +
    geom_line()+
    xlab("Sequencing Depth")+
    ylab("Number of ASVs")+
    ggtitle("Rarefaction plot, 1185mR, Maize samples") +
    plot_theme
```

The rarefaction curves show a higher absolute diversity for the 1185mR primer pair but in general over the samples it's the opposite. 

```{r}
#| fig-height: 12
#| fig-width: 10

  physeq.1185.filtered <- prune_taxa(rownames(tax.bac.1185), physeq.1185)
  physeq.1185.phylum <- tax_glom(physeq.1185.filtered, taxrank = "Phylum")

  physeq.1185.phylum.rel <- transform_sample_counts(
    physeq.1185.phylum,
    function(x) x / sum(x)
  )

  percentages_1185.rel.loic <- psmelt(physeq.1185.phylum.rel) %>% filter(SampleOwner == "Loïc")
  
  
  relative_plot <- ggplot(data=percentages_1185.rel.loic, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1185mR") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
  
  physeq.1492.filtered <- prune_taxa(rownames(tax.bac.1492), physeq.1492)
  physeq.1492.phylum <- tax_glom(physeq.1492.filtered, taxrank = "Phylum")

  physeq.1492.phylum.rel <- transform_sample_counts(
    physeq.1492.phylum,
    function(x) x / sum(x)
  )

  percentages_1492.rel.loic <- psmelt(physeq.1492.phylum.rel) %>% filter(SampleOwner == "Loïc")
  
  
  relative_plot <- ggplot(data=percentages_1492.rel.loic, aes(x=sample_Sample, y=Abundance, fill=Phylum))+
    geom_bar(aes(), stat="identity", position="stack") +
    theme_bw() + facet_wrap(~host) + ggtitle("1492R") + plot_theme +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90)
    ) 
  
  relative_plot
```








